VOI-aware MCTS

### 2012-02-26 paper/vago.ecai/paper.tex
short paper underway

\title{VOI-aware MCTS}
\author {David Tolpin, Solomon Eyal Shimony \\
Department of Computer Science, \\
Ben-Gurion University of the Negev, Beer Sheva, Israel \\
\{tolpin,shimony\}@cs.bgu.ac.il}

@@

\begin{dfn} The \textbf{blinkered estimate} of intrinsic VOI information of the
  $i$th arm is the intrinsic VOI of pulling the $i$th arm for the
  remaining budget.
\end{dfn}

\begin{thm} Denote the current number of samples of the $i$th arm by
  $n_i$. The blinkered estimate $\Lambda_i^b$ of intrinsic value of
  information of pulling the $i$th arm for the remaining budget of $N$
  samples is bounded from above as
\begin{equation}
  \Lambda_i^b \le \left\{
  \begin{array}{l l}
    \overline X_\beta\frac N {N+n_i}\Pr(\overline X_i'\le\overline X_\beta)
    \le N\frac {\overline X_\beta} {n_i}\Pr(\overline X_i'\le\overline X_\beta) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\frac N {N+n_i}\Pr(\overline X_i'\ge\overline X_\alpha)
       \le N\frac {(1-\overline  X_\alpha)} {n_i}\Pr(\overline X_i'\ge\overline X_\alpha) & \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:thm-be}
\end{equation}
where $\overline X_i'$ is the sample mean of the $i$th arm after $n_i+N$ 
samples.
\label{thm:be}
\end{thm}

@@

The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be})
can be bounded from above using concentration inequalities. In
particular, Theorem~\ref{thm:hoeffding-prob-bounds} (proved in 
Appendix~\ref{app:hoeffding-prob-bounds-proof}) is based on the
Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{thm} The probabilities in equations (\ref{eqn:thm-vopi},
\ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)&
   \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)&
   \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\vspace{-1.5\baselineskip}
\begin{eqnarray}
\Pr(\overline X_i' \le \overline X_\beta|i=\alpha) &
   \le & 2\exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}}(\overline X_i - \overline X_\beta)\right)^2 n_i \right)
   \le 2\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i \right) \nonumber\\
\Pr(\overline X_i' \ge \overline X_\alpha|i\ne\alpha) &
   \le & 2\exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}} (\overline X_\alpha - \overline X_i)\right)^2 n_i \right)
    \le 2\exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i \right)
\label{eqn:probound-blnk-hoeffding}
\end{eqnarray}
\label{thm:hoeffding-prob-bounds}
\end{thm}

\begin{crl} An upper bound on the intrinsic value of perfect information is obtained by substituting
(\ref{eqn:probound-perf-hoeffding}) into (\ref{eqn:thm-vopi}):
\begin{equation}
  \Lambda_i^p \le \hat\Lambda_i^p=\left\{
  \begin{array}{l l}
    \overline X_\beta\cdot \exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\cdot \exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i \right)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-perf-hoeffding}
\end{equation}
\label{crl:bound-perf-hoeffding}
\end{crl}
\begin{crl}
An upper bound on the blinkered VOI estimate with ordering independent of $N$ is obtained
by substituting (\ref{eqn:probound-blnk-hoeffding}) into (\ref{eqn:thm-be}).
\begin{equation}
  \Lambda_i^b \le \hat\Lambda_i^b=\left\{
  \begin{array}{l l}
    N\frac {\overline X_\beta} {n_i}\cdot 2\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    N\frac {(1-\overline  X_\alpha)} {n_i}\cdot 2\exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-blnk-hoeffding}
\end{equation}
\label{crl:bound-blnk-hoeffding}
\end{crl}
Bound~\ref{eqn:bound-blnk-hoeffding} is generally tighter when $N$ is
small relative to $n_i$. Estimate $\hat\Lambda_i^b$ can be viewed as a
relaxation of the myopic assumption such that sequences consisting
of more than one action are taken into considertion. Better bounds can
be obtained through tighter estimates on the probabilities, for
example, based on the empirical Bernstein inequality
(\ref{eqn:conc-empbernstein}) or through a more careful application of
the Hoeffding inequality (Appendix~\ref{app:better-hoeffding-bound}).

@@

The experiments compare the UCT algorithm \cite{Kocsis.uct} with
modified versions of UCT in which the samples at the first step are
selected according to their VOI estimates. Three blinkered VOI
estimates were used for the comparison:
\begin{description}
\item[VCT:] based on Hoeffding inequality (\ref{eqn:conc-hoeffding});
\item[ECT:] based on Hoeffding inequality with an intermediate point (see
Appendix~\ref{app:better-hoeffding-bound});
\item[BCT:] based on empirical Bernstein inequality
(\ref{eqn:conc-empbernstein}).
\end{description}
For simplicity, the bounds on probability for an unlimited number
of samples were used in the estimate formulas. The estimates where
computed as follows:
\begin{eqnarray}\
  \Lambda_i^{VCT}& =&\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta} {n_i} \cdot \exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    \frac {(1-\overline  X_\alpha)} {n_i}\cdot \exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array}  \right.\nonumber\\
  \Lambda_i^{ECT}& =&\min\left(\Lambda_i^{VCT}, \min_y\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta-y} {n_i}\cdot \exp\left(-2 (\overline X_i-\overline X_\beta)^2n_i\right)
    +\frac {\overline X_\beta} {n_i}\cdot \exp\left(-2 (\overline X_i-y)^2n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
\frac {y-\overline X_\alpha} {n_i}\cdot \exp\left(-2(\overline X_\alpha-\overline X_i)^2n_i\right)
    +\frac {1-\overline X_\alpha} {n_i}\cdot \exp\left(-2 (y-\overline X_i)^2n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.\right) \nonumber\\
  \Lambda_i^{BCT}& =&\min\left(\Lambda_i^{VCT},\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta} {n_i}\cdot 2\exp\left(- \frac {(\overline X_i - \overline X_\beta)^2 n_i} {\frac {14} 3 \frac {n_i} {n_i-1}(\overline X_i - \overline X_\beta) + 2 \overline X_i(1-\overline X_i)} \right)
      & \mbox{{\rm if} $i=\alpha$} \\
    \frac {(1-\overline  X_\alpha)} {n_i}\cdot 2\exp\left(- \frac {(\overline X_\alpha - \overline X_i)^2 n_i} {\frac {14} 3 \frac {n_i} {n_i-1}(\overline X_\alpha - \overline X_i) + 2 \overline X_i(1-\overline X_i)} \right)
      &  \mbox{\rm otherwise}
  \end{array} \right.\right)
\label{eqn:emp-estimates}
\end{eqnarray}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{vct-wins.pdf}
\caption{Go: winning rate --- UCT against VCT}
\label{fig:uct-against-vct}
\end{figure}
\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{ect-wins.pdf}
\caption{Go: winning rate --- UCT against ECT}
\label{fig:uct-against-ect}
\end{figure}
\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{bct-wins.pdf}
\caption{Go: winning rate --- UCT against BCT}
\label{fig:uct-against-bct}
\end{figure}
The engines were compared on the 9x9 board, for 5000, 7000, 1000, and
15000 samples (game simulations) per ply, each experiment was repeated
1000 times. Figures~\ref{fig:uct-against-vct}--\ref{fig:uct-against-bct}
show the winning rate of UCT against VCT, ECT, BCT, correspondingly,
vs. the stopping threshold (if the maximum VOI of a sample is below
the threshold, the simulation is stopped, and a move is chosen). Each
curve in the figures corresponds to a certain number of samples per
ply. The lower the curve, the better is the corresponding policy
compared to UCT.

As the results show, without VOI-aware sample redistribution the
VOI-aware policies are at best equal to UCT for lower numbers of
samples per ply (5000, 7000), and the advantage of the VOI-aware policies
for larger number of samples is rather modest (e.g., ECT wins in 55\%
of games for 10000 samples per ply). For the stopping threshold of
$10^{-6}$, however, the VOI-aware policies are almost always better
than UCT (only VCT for 5000 samples wins in less than 50\% of games),
with stronger policies (ECT, VCT) reaching the winning rate of 66\%.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{bests-bw.pdf}
\caption{Go: best winning rate comparison}
\label{fig:best-winning-rate}
\end{figure}
Figure~\ref{fig:best-winning-rate} presents relative performance of
different VOI-aware policies depending on the number of samples per
ply by comparing the best achieved winning rates against UCT (again,
the lower the bar, the better is the policy). Both ECT (based on the
Hoeffding inequality with an intermediate point) and BCT (based on the empirical
Bernstein inequality) are better than VCT. The difference between ECT
and VCT is most prominent for smaller number of samples, and the
winning rate of ECT increases relatively slowly with the number of
samples. In contrast, BCT is worse than ECT and only slightly better
than VCT for smaller number of samples, but the winning rate rapidly
increases with the number of samples, and for 10000 and 15000 samples
per ply BCT is the best of the policies with the winning rate reaching
67\%.

@@

\begin{proof}[]
Equations (\ref{eqn:probound-perf-hoeffding}) is a direct
application of the Hoeffding inequality (\ref{eqn:conc-hoeffding}).

>>>>

\title{VOI-aware MCTS}
%\author {David Tolpin, Solomon Eyal Shimony \\
%Department of Computer Science, \\
%Ben-Gurion University of the Negev, Beer Sheva, Israel \\
%\{tolpin,shimony\}@cs.bgu.ac.il}

@@

\begin{thm} Denote the current number of samples of the $i$th arm by
  $n_i$. The intrinsic value of
  information $\Lambda_i^b$ of pulling the $i$th arm for the remaining budget of $N$
  samples is bounded from above as
\begin{equation}
  \Lambda_i^b \le \left\{
  \begin{array}{l l}
    \overline X_\beta\frac N {N+n_i}\Pr(\overline X_i'\le\overline X_\beta)
    \le N\frac {\overline X_\beta} {n_i}\Pr(\overline X_i'\le\overline X_\beta) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\frac N {N+n_i}\Pr(\overline X_i'\ge\overline X_\alpha)
       \le N\frac {(1-\overline  X_\alpha)} {n_i}\Pr(\overline X_i'\ge\overline X_\alpha) & \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:thm-be}
\end{equation}
where $\overline X_i'$ is the sample mean of the $i$th arm after $n_i+N$ 
samples.
\label{thm:be}
\end{thm}

@@

The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be})
can be bounded from above using concentration inequalities. In
particular, Theorem~\ref{thm:hoeffding-prob-bounds} (proved in 
Appendix~\ref{app:hoeffding-prob-bounds-proof}) is based on the
Hoeffding inequality \cite{Hoeffding.ineq}.
\begin{thm} The probabilities in equations (\ref{eqn:thm-vopi},
\ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)&
   \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)&
   \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\vspace{-1.5\baselineskip}
\begin{eqnarray}
\Pr(\overline X_i' \le \overline X_\beta|i=\alpha) &
   \le & 2\exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}}(\overline X_i - \overline X_\beta)\right)^2 n_i \right)
   \le 2\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i \right) \nonumber\\
\Pr(\overline X_i' \ge \overline X_\alpha|i\ne\alpha) &
   \le & 2\exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}} (\overline X_\alpha - \overline X_i)\right)^2 n_i \right)
    \le 2\exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i \right)
\label{eqn:probound-blnk-hoeffding}
\end{eqnarray}
\label{thm:hoeffding-prob-bounds}
\end{thm}

\begin{crl} An upper bound on the intrinsic value of perfect information is obtained by substituting
(\ref{eqn:probound-perf-hoeffding}) into (\ref{eqn:thm-vopi}):
\begin{equation}
  \Lambda_i^p \le \hat\Lambda_i^p=\left\{
  \begin{array}{l l}
    \overline X_\beta\cdot \exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\cdot \exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i \right)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-perf-hoeffding}
\end{equation}
\label{crl:bound-perf-hoeffding}
\end{crl}
\begin{crl}
An upper bound on the blinkered VOI estimate with ordering independent of $N$ is obtained
by substituting (\ref{eqn:probound-blnk-hoeffding}) into (\ref{eqn:thm-be}).
\begin{equation}
  \Lambda_i^b \le \hat\Lambda_i^b=\left\{
  \begin{array}{l l}
    N\frac {\overline X_\beta} {n_i}\cdot 2\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    N\frac {(1-\overline  X_\alpha)} {n_i}\cdot 2\exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-blnk-hoeffding}
\end{equation}
\label{crl:bound-blnk-hoeffding}
\end{crl}
Bound~\ref{eqn:bound-blnk-hoeffding} is generally tighter when $N$ is
small relative to $n_i$. Estimate $\hat\Lambda_i^b$ can be viewed as a
relaxation of the myopic assumption such that sequences consisting
of more than one action are taken into considertion. Better bounds can
be obtained through tighter estimates on the probabilities, for
example, based on the empirical Bernstein inequality
 or through a more careful application of
the Hoeffding inequality.

@@

The experiments compare the UCT algorithm \cite{Kocsis.uct} with
a modified version of UCT in which the samples at the first step are
selected according to their VOI estimates.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{vct-wins.pdf}
\caption{Go: winning rate --- UCT against VCT}
\label{fig:uct-against-vct}
\end{figure}

The engines were compared on the 9x9 board, for 5000, 7000, 1000, and
15000 samples (game simulations) per ply, each experiment was repeated
1000 times. Figure~\ref{fig:uct-against-vct}
show the winning rate of UCT against VCT
vs. the stopping threshold (if the maximum VOI of a sample is below
the threshold, the simulation is stopped, and a move is chosen). Each
curve in the figures corresponds to a certain number of samples per
ply. The lower the curve, the better is the corresponding policy
compared to UCT.

@@

@@

\begin{proof}[]
Equations (\ref{eqn:probound-perf-hoeffding}) is a direct
application of the Hoeffding inequality \cite{Hoeffding.ineq}.
